01_Vegetation_cover

### 2022-11-14 Manuscript.Rmd
update after comments

# Reconstructing plant trait composition from palaeoecological data: testing the validity of the use of pollen data for functional trait reconstruction

@@

Functional traits provide a better insight in to the processes that drive biodiversity and how biodiversity affects ecosystem functioning [@Hevia_2017]. While the functioning of modern ecosystems is regularly contingent on past human and environmental factors[@Levis_2018; @Jogiste_2017; @Blonder_2018], the observational duration of trait-based studies is generally short[@Estes_218]. So to extend the time scales of ecosystem functioning research, considerable interest in the trait-based approach has arisen in palaeoecology [@Brussel_2018; @Blaus_2020; @Carvalho_2019; @Reitalu_2015; @Sande_2019]. Applying the trait-based approach in palaeoecological studies could offer a new way of interpreting pollen data and increase our understanding of the drivers of vegetation change on timescales of centuries to millennia. For instance, @Sande_2019 found that on a 7000-year time scale human disturbance superseeds climate effects on plant trait composition in the Amazon.

Although palaeoecological studies that have applied the trait-based approach show promising results, it is a relatively new approach of which the intricacies are underexplored. Little is known about how representative the pollen record is of past plant trait composition. Two main methodological issues arise when attempting to reconstruct functional trait composition from pollen data. First, pollen quantity does not exactly represent past plant abundance. Second, the low taxonomic resolution of the pollen record makes it not straightforward to assign traits to pollen taxa, as trait data is generally collected at the species level.

Taxonomic differences in pollen dispersal and production can cause an over- or underrepresentation of taxa in the pollen record. Wind-pollinated and woody taxa are generally better represented in the pollen record than herbaceous and insect-pollinated taxa as their pollen are produced in larger quantities and dispersed over greater distances. The pollen source area for wind pollinated taxa can therefore be a couple magnitudes greater than that of wind pollinated taxa [@Meltsov_2011] (extend argument? bunting 2003, bunting 2022). It has therefore been suggested by @Blaus_2020 to treat woody and non-woody taxa separately in the reconstruction of functional composition from pollen data. Another or additional way to correct for taxon specific differences use pollen productivity estimates, these correction factors down-weight the contribution of dominant taxa [@Githumbi_2022].

Perhaps the most challenging in reliably reconstructing past trait is to attribute functional trait data to the taxa found in pollen records. It is rarely possible to resolve the pollen record to the species level [@Birks_2020]. Trait variation within taxa can be substantial depending on the trait and the diversity of the taxon [@Albert_2010; @Prinzing_2008]. It can be expected that this especially problematic in the case of large, abundant plant families that cannot easily be resolved to a lower taxonomic level, such as the Poaceae family. @Blaus_2020 and @Carvalho_2019 proposed that the high species number included in herbaceous pollen taxon could be one of the reasons why herbaceous vegetation was less well represented by pollen based functional diversity reconstructions. In both studies, however, traits values of pollen taxa were calculated by averaging over the available species level trait information. In this study, we use a Bayesian approach for modelling trait composition from pollen data, thereby allowing for the propagation of uncertainty in the taxon level trait reconstruction to the community level.

With this study, we aim to test the validity of using of pollen records for plant trait composition reconstruction. We examine the effectiveness of pollen production factors to improve the pollen - functional diversity relationship. Furthermore, we test the representation of different plant functional groups by the pollen-based functional trait reconstruction. Laslty, we investigate the effect of taxonomic resolution on the reconstructions.

@@

We collected pollen, trait and species composition data from 16 sites in the Cairngorms National Park, Scotland, United Kingdom. Sites were selected to represent four vegetation types in the area: grassland, heath land, deciduous forest and coniferous forest. Sampling locations were at least 1 km from each other to avoid pollen source areas from overlapping with each other. Pollen, trait and species data was also available from `r length(bdm_meta)` sites across Switzerland were monitored in the context of the Biodiversity Monitoring Switzerland (BDM). This is a systematic sampling grid for the long-term monitoring of plant and animal species in Switzerland [@BDM_2022]. Of these sites `r table(bdm_meta$habitat01)[1]` were forest and `table(bdm_meta$habitat01)[2]` grassland, spanning an elevation between `r min(bdm_meta$elevation)` and `r max(bdm_meta$elevation)`. Trait and species composition data from `r nrow(read.csv("Data/Swiss_LC/all_BDM_plots_coord.csv")) - length(bdm_meta)` were also used in this study.

@@

We compare trait composition of vegetation with trait composition reconstructed from surface pollen samples. Surface pollen samples are collected from moss polsters and approximate vegetation composition in recent years [@Birks_2016] [Other ref?]. Study design mostly followed the protocol of @Bunting_2013 (Appendix), which allows for the calculation of vegetation composition in concentric circles around the moss sampling point. We calculated the vegetation composition in three nested concentric rings of 3.4m, 100m and 1000m in radius. The inner ring (3.4m) was surveyed in the field, the middle ring (100m) was surveyed in Scotland, but not in Switzerland. The vegetation cover of the outer ring (1000m) and the middle ring for Switzerland was estimated using QGIS and land cover maps [@SwissFC; @SwissLC;\@scotisch_government;\@QGIS_software].

### Vegetation data

In the inner ring (3.4m) 4 plots of 0.25 m^2^ (0.16m^2^ in Switzerland) were surveyed, one in every wind direction. At the Scottisch sites the middle ring was surveyed as well, between 4 and 6 plots of 10 by 10 meter were selected on a stratified random design based on vegetation types that were identified in the fields. Species abundance was estimated using percentage cover in Scotland. In Switzerland, each subplot was a quadrat with nine points (a 3 x 3 grid), and percentage coverage of the species present at the nine points was estimated. All species names were standardized according to the Leipzig Plant Catalogue using the *lcvplants* and *lcvp* R packages [@freiberg2020lcvp].

Three traits were selected for the analysis, specific leaf area, leaf area and and plant height. These traits reflect different components in the spectrum of plant strategies. XXX. In Scotland, trait data was collected in the middle ring (100m). For determining plant height 10 flowering of individuals were selected at every site if present, non-flowering individuals were selected if flowering individuals were not available. Five leaf samples per species for specific leaf area (SLA), leaf area (LA), leaf dry matter content (LDMC), leaf carbon and leaf nitrogen were collected in seal-able plastic bags. Fresh weight and surface area were measured within 72 hours of collection using a carat scale and flatbed scanner. Samples were dried at 70 C for 72 hours to constant dry weight before measuring dry weight. In Switzerland, the trait data was collected for species present at the nine points in the 3 by 3 grid.

For reconstructing trait composition from pollen data and for classifying vegetation data in to plant growth forms and pollination mode, trait data from the TRY database was used [@kattge2020try] (Appendix).

@@

Moss was collected from multiple location within a 1m radius in the centre of the inner ring. Only living tissue was sampled. Pollen were extracted from the moss using standardized pollen preparation protocol (REF). Pollen were identified using light microscope and the key of (REF). To test the ability of pollen correction factors to improve pollen - vegetation relationships, pollen productivity estimates from @Githumbi_2022 were used. Pollen productivity estimates were not available for all taxa, taxa without pollen productivity factors were excluded. On average, this led to the exclusion of `r round(mean(missing_ppe_scot$percent_missing),2)*100`% of the Scottish pollen data (max `r round(max(missing_ppe_scot$percent_missing),2)*100`%) and `round(mean(missing_ppe_swiss$percent_missing),2)*100`% of the Swiss pollen data (max `r round(max(missing_ppe_swiss$percent_missing),2)*100`%). Pollen were divided in to two plant functional groups according to @Reitalu_2019 consisting of woody (trees and shrubs) ornon-woody species (grasses, herbs and dwarf shrubs of the Ericaceae family). @Reitalu_2019 was also used to classify pollen in to pollination mode, either insect-pollinated or not insect-pollinated.

For reconstructing plant trait composition from pollen data, assumptions need to be made about the species belonging to the pollen taxa in the data set. We used the *rgbif* package to download the current distribution of species belonging to each pollen taxon in Scotland and Switzerland from the Global Biodiversity Information Facility database (REF). Species distribution data was queried using the pollen taxonomic name, i.e. the lowest taxonomic name the pollen could be identified to .

@@

We use the community weighted mean (CWM) to characterize functional composition, which here is specified as the trait distribution of a taxon, i.e. the species in the vegetation data and the pollen taxon in the pollen data, weighted by its abundance. First, the trait distribution on the species or the pollen taxonomic level was estimated. We assume that the trait data follows a lognormal distribution for the likelihood, so that$$trait_{i}^{observed} \sim { lognormal(\mu_{j},\sigma_j)}$$Where *i* represents each trait observation of a taxon (*j*) from the trait data. $\mu_{j}$ and $\sigma_{j}$ are the mean trait value and the standard deviation per taxon, respectively. Vague priors were used for the taxon mean and standard deviation estimation. We then modeled the community weighted mean (CWM) as a mixture of normal distributions that were weighted by the species abundance or pollen percentage, so that

@@

Where *n* is the total number of taxa, and $\pi_{j}$ the abundance of each pollen taxon or species at a given site *s*. The mean *CWM~­~^mean^* and the standard deviation *CWM^sd^* of this modeled distribution was subsequently used in the estimation of the correlation between the trait composition of the vegetation and the pollen data.

### Comparison between pollen and vegetation community weighted mean

To test the correlation between the pollen based reconstruction of CWM and the vegetation CWM, we estimated the correlation coefficient *ρ* we assumed the data follows a multivariate t likelihood distribution so that, $$[CWM_{s}^{pol},CWM_{s}^{veg}] \sim t_{df}([\mu_{pol},\mu_{veg}],\sum) $$

Where we estimated the degrees of freedom ($df$) of the multivariate t-distribution with and exponential prior with a mean of 30 as the t-distribution becomes almost normal at values of 30 and above (Kruscke, dbda),

$$
df \sim Exp(\lambda)
$$

$$
\lambda = 1/30
$$

We used an uninformative, flat prior for $\rho$ so that,

@@

Correlations were calculated for all zones and three traits: SLA, LA and plant height. Furthermore, four different treatments were assigned to test what factors could influence representation of trait composition by pollen data: the use of correction factors, the separation of vegetation and pollen data in to plant functional type and pollination mode. Lastly, the vegetation was aggregated to genus level and family level to test the effect of taxonomic resolution on the reconstructions.

@@

@@

@@

@@

@@

@@

```{r echo=FALSE, fig.height=6, fig.width=10}
source("03_Vegetation_pollen_comparison.R") # pollen plot
pol_veg_plot
```

```{r echo=FALSE, fig.height=6, fig.width=10}
source("03_Vegetation_pollen_comparison.R") # pollen plot
pol_veg_plot_adj
```

### Site characteristics

insert pollen plot

```{r echo=FALSE}
# source("03_pollen_diagrams.R")
```

@@

![Design of Swiss trait sampling and vegetation surveys](Other/BDM_study_design.png)

![Design of Scottish vegetation survey in zone A](Other/Scotland_field_design.png)

### Trait values

```{r echo=FALSE}
sum_trait <- readRDS("RDS_files/01_summary_table_TRY_data.rds")
kable(sum_trait)
```

```{r echo=FALSE}
source("01_Prepare_trait_data_vegetation.R",encoding ="UTF-8")
kable(sum_ph)
kable(sum_la)
kable(sum_sla)
```

>>>>

# Reconstructing plant trait composition from palaeoecological data: testing the validity of the use of pollen data for studying long-term changes in plant trait composition

@@

Functional traits provide a better insight in to the processes that drive biodiversity and how biodiversity affects ecosystem functioning [@Hevia_2017]. While the functioning of modern ecosystems is regularly contingent on past human and environmental factors [@Levis_2018; @Jogiste_2017; @Blonder_2018; ], the observational duration of trait-based studies is generally at most a couple of decades [@Estes_218]. So to extend the time scales of ecosystem functioning research, considerable interest in the trait-based approach has arisen in palaeoecology [@Brussel_2018; @Blaus_2020; @Carvalho_2019; @Reitalu_2015; @Sande_2019]. Applying the trait-based approach in palaeoecological studies could offer a new way of interpreting pollen data and increase our understanding of the drivers of vegetation change on time-scales of centuries to millennia. For instance, @Sande_2019 found that on a 7000-year time scale human disturbance supersedes climate effects on plant trait composition in the Amazon.

Although palaeoecological studies that have applied the trait-based approach show promising results, it is a relatively new approach of which the intricacies are under-explored. Little is known about how representative the pollen record is of past plant trait composition. Two main methodological issues arise when attempting to reconstruct functional trait composition from pollen data. First, pollen quantity does not exactly represent past plant abundance. Second, the low taxonomic resolution of the pollen record makes it not straightforward to assign traits to pollen taxa, as trait data is generally collected at the species level.

Taxonomic differences in pollen dispersal and production can cause an over- or under-representation of taxa in the pollen record. Wind-pollinated and woody taxa are generally better represented in the pollen record than herbaceous and insect-pollinated taxa as their pollen are produced in larger quantities and dispersed over greater distances. The source area of a pollen of a wind-pollinated taxon can therefore be a couple magnitudes greater than that of a insect-pollinated taxon [@Meltsov_2011] (extend argument? bunting 2003, bunting 2022). Over- or under-representation can be corrected for with the use of taxon-specific pollen productivity estimates, these correction factors down-weight the contribution of dominant taxa [@Githumbi_2022]. Furthermore, it has been suggested by @Blaus_2020 to treat woody and non-woody taxa separately in the reconstruction of functional composition from pollen data as their representative pollen source area differs greatly.

Perhaps the most challenging in reliably reconstructing past trait composition is to attribute functional trait data to the taxa found in pollen records, as it is rarely possible to resolve the pollen record to the species level [@Birks_2020]. Trait variation within taxa can be substantial depending on the trait and the diversity of the taxon [@Albert_2010; @Prinzing_2008]. It can be expected that this is especially problematic in the case of large, abundant plant families that cannot easily be resolved to a lower taxonomic level, such as the Poaceae family. @Blaus_2020 and @Carvalho_2019 proposed that the high species number included in herbaceous pollen taxon could be one of the reasons why herbaceous vegetation was less well represented by pollen based functional diversity reconstructions. In both studies, however, traits values of pollen taxa were calculated by averaging over the available species level trait information. In this study, we use a Bayesian approach for modelling trait composition from pollen data, thereby allowing for the propagation of uncertainty in the taxon level trait reconstruction to the community level.

With this study, we aim to test the validity of using pollen records for past plant trait composition reconstruction. We address two questions. Firstly, what factors influence the pollen - vegetation trait relationship? Using a data set from the Cairngorms, Scotland, we investigate the effect of spatial scale, plant functional group, correction factors and taxonomic resolution on the representativeness of the pollen assemblages for functional trait reconstruction. Secondly, can pollen based trait reconstruction predict changes in trait composition along environmental gradients? Using pollen, vegetation and trait data collected along an elevation gradient in Switzerland we test if pollen assemblages can predict changes in trait composition with elevation.

@@

We collected pollen, trait and species composition data from 16 sites in the Cairngorms National Park, Scotland, United Kingdom. Sites were selected to represent four vegetation types in the area: grassland, heath land, deciduous forest and coniferous forest. Sampling locations were at least 1 km from each other to avoid pollen source areas from overlapping with each other. Pollen, trait and species data was also available from `r nrow(bdm_meta)` sites across Switzerland were monitored in the context of the Biodiversity Monitoring Switzerland (BDM). This is a systematic sampling grid for the long-term monitoring of plant and animal species in Switzerland [@BDM_2022]. Trait and species composition data from an additional `r nrow(read.csv("Data/Swiss_LC/all_BDM_plots_coord.csv")) - nrow(bdm_meta)` sites that were surveyed in this network were also used in this study.

@@

The data that was collected in Scotland was used to test if pollen-based trait reconstructions represents vegetation trait composition. We compare the community weighted mean (CWM) trait value of the vegetation with the CWM trait values reconstructed from surface pollen samples. Surface pollen samples are collected from moss polsters and approximate vegetation composition in recent years [@Birks_2016] [+Other ref?]. Study design mostly followed the protocol of @Bunting_2013, which allows for the calculation of vegetation composition in concentric circles around the moss sampling point to test the spatial scale that is represented by the moss sample (Appendix). We calculated the vegetation composition in three nested concentric rings of 10 m, 100 m and 1000 m in radius. The inner ring (10 m) and the middle ring was surveyed in the field. The vegetation cover of the outer ring (1000 m) was estimated using QGIS and land cover maps [\@scotisch_government; \@QGIS_software]. To test the effect of spatial scale on the representativeness of pollen assemblages for trait composition reconstruction, we test compare vegetation CWM and pollen CWM in the rings of 10 m, 100 m and 1000 m. Furthermore, four different treatments were assigned to test what factors could influence representation of trait composition by pollen data. Firstly, the use of correction factors, than the separation of vegetation and pollen data in to plant functional type and pollination mode. Lastly, the vegetation was aggregated to genus level and family level to test the effect of taxonomic resolution on the reconstructions.

To test if pollen-trait reconstructions can predict relative changes in vegetation trait composition along environmental gradients, we calculate changes in CWM trait values along an elevation gradient in Switzerland. The sites span an elevation between `r min(bdm_meta$elevation)` m and `r max(bdm_meta$elevation)`m. Of these sites `r table(bdm_meta$habitat01)[1]` were in forests and `r table(bdm_meta$habitat01)[2]` in grassland.

### Vegetation survey

At the Scottish sites, in the inner ring (10 m) 17 sub-plots of 0.25 m^2^ were surveyed: one in the centre, 3 in every wind direction at 3, 6 and 10 meters, and one in the NE, NW, SE and SW direction (See design in Appendix). In the middle ring (10-100 m) between 4 and 6 plots of 5 by 5 meter were selected on a stratified random design based on vegetation types that were identified in the fields. Vegetation abundance was estimated using percentage cover. In Switzerland, the plots were 1.8 m in radius and contained 4 sub-plots. Each sub-plot was a quadrant of 0.16 m^2^ with nine points (a 3 x 3 grid), and percentage coverage of the species present at the nine points was estimated. All species names were standardized according to the Leipzig Plant Catalogue using the *lcvplants* and *lcvp* R packages [@freiberg2020lcvp].

### Trait data collection

Three traits were selected for the analysis, specific leaf area, leaf area and and plant height. These traits reflect different components in the spectrum of plant strategies. XXX. In Scotland, trait data was collected in the middle ring (100 m). For determining plant height 10 flowering of individuals were selected at every site if present, non-flowering individuals were selected if flowering individuals were not available. Five leaf samples per species for specific leaf area (SLA) and leaf area (LA) collected in seal-able plastic bags. Fresh weight and surface area were measured within 72 hours of collection using a carat scale and flatbed scanner. Samples were dried at 70 ***°***C for 72 hours to constant dry weight before measuring dry weight. In Switzerland, the trait data was collected for species present at the nine points in the 3 by 3 grid. XXXX

For reconstructing trait composition from pollen data and for classifying vegetation data in to plant growth forms and pollination mode, previously gapfilled trait data from the TRY database was used [@kattge2020try] [\@Schrodt_2015]. XXX

@@

Moss was collected from multiple location within a 1 m radius in the centre of each plot. Only living tissue was sampled. Pollen were extracted from the moss using standardized pollen preparation protocol (REF). Pollen were identified using light microscope and the key of (REF). To test the ability of pollen correction factors to improve pollen - vegetation relationships, pollen productivity estimates from @Githumbi_2022 were used. Pollen productivity estimates were not available for all taxa, taxa without pollen productivity factors were excluded. On average, this led to the exclusion of `r round(mean(missing_ppe_scot$percent_missing),2)*100`% of the Scottish pollen data (max `r round(max(missing_ppe_scot$percent_missing),2)*100`%). Pollen were divided in to two plant functional groups according to @Reitalu_2019 consisting of woody (trees and shrubs) or non-woody species (grasses, herbs and dwarf shrubs of the Ericaceae family). @Reitalu_2019 was also used to classify pollen in to pollination mode, either insect-pollinated or not insect-pollinated.

For reconstructing plant trait composition from pollen data, assumptions needed to be made about the species belonging to the pollen taxa in the data set. Therefore we created a pollen to species translation table. We used the *rgbif* package to download the current distribution of species belonging to each pollen taxon in Scotland and Switzerland from the Global Biodiversity Information Facility database (REF). Species distribution data was queried using the pollen taxonomic name, i.e. the lowest taxonomic name the pollen could be identified to. All trait data that was available for the species in the translation table was then used in the calculation trait values on the pollen taxonomic level .

@@

We use the community weighted mean (CWM) to characterize functional composition, which here is specified as the trait distribution of a taxon, i.e. the species in the vegetation data and the pollen taxon in the pollen data, weighted by its abundance. First, the trait distribution on the species or the pollen taxonomic level was estimated. We assume that the trait data of SLA, LA and plant height follow a lognormal distribution for the likelihood, so that$$trait_{i}^{observed} \sim { lognormal(\mu_{j},\sigma_j)}$$Where *i* represents each trait observation of a taxon (*j*) from the trait data. $\mu_{j}$ and $\sigma_{j}$ are the mean trait value and the standard deviation per taxon, respectively. Vague priors were used for the taxon mean and standard deviation estimation. We then modelled the community weighted mean (CWM) as a mixture of normal distributions that were weighted by the species abundance or pollen percentage, so that

@@

Where *n* is the total number of taxa, and $\pi_{j}$ the abundance of each pollen taxon or species at a given site *s*. The mean *CWM~­~^mean^* and the standard deviation *CWM^sd^* of this modelled distribution was subsequently used in the estimation of the correlation between the trait composition of the vegetation and the pollen data.

### Correlation between pollen and vegetation community weighted mean

To test the correlation between the pollen based reconstruction of CWM and the vegetation CWM, we used a multivariate normal distribution to estimate correlation coefficient $\rho$ so that,

$$  [CWM_{s}^{veg},CWM_{s}^{veg}] \sim MultivariateNormal([\mu_{veg},\mu_{veg}],\sum) $$

@@

Correlations were calculated for the different treatments, all zones and three traits: SLA, LA and plant height.

### Community weighted mean changes with elevation

To test if relative trait compositional changes are reflected by the pollen data along an elevation gradient, we fitted a generalized linear model with elevation as the predictor, so that

$$CWM^{mean} \sim Normal(\alpha + \beta Y_{elev}, CWM^{sd})$$

Where $\alpha$ and $\beta$ are the intercept and the slope, respectively, which are supplied by uninformative priors. $Y_{elev}$ is the elevation as the linear predictor.

@@

![Estimated correlation between the pollen data and vegetation data. Whisker plots show the median estimated correlation coefficient and the 95% credibility interval.](Figures/Pollen_correction_factors.png)

\* add uncertainty around pollen correction factors

@@

![Estimated correlation between the pollen data and vegetation data. Whisker plots show the median estimated correlation coefficient and the 95% credibility interval.](Figures/Pollination_mode.png)

@@

![Estimated correlation between the pollen data and vegetation data. Whisker plots show the median estimated correlation coefficient and the 95% credibility interval.](Figures/PFT.png)

@@

![Estimated correlation between the pollen data and vegetation data. Whisker plots show the median estimated correlation coefficient and the 95% credibility interval.](Figures/Taxonomic_resolution.png)

### CWM change with elevation

![CWM trait change with elevation for Leaf area, plant height and specific leaf area. Left column shows the CWM values for the vegetation data, right column shows the CWM values for the pollen data. Green lines show the relationships from the linear model, grey area shows the 95% credibility interval. Leaf area and plant height were log transformed before modelling.](Figures/20221016-Swiss_elevation_combined.png)

@@

-   Interspecific trait variation in LA

@@

![Scatter plot of vegetation basal area % against pollen abundance %](Figures/Pollen_rep_percent.png)

![Scatter plot of vegetation basal area % against pollen abundance % adjusted with pollen productivity estimates.](Figures/Pollen_rep_adjustedpercent.png)

@@

![Study design in Scotland. The moss sample was collected in the middle of the inner ring. Vegetation surveys were performed in the inner and middle ring. Trait data was collected in the middle ring. The outer ring was surveyed using land cover maps and QGIS](images/paste-941650FE.png){alt="Study design in Scotland. The moss sample was collected in the middle of the inner ring. Vegetation surveys were performed in the inner and middle ring. Trait data was collected in the middle ring. The outer ring was surveyed using land cover maps and QGIS" width="573"}

![Design of Scottish vegetation survey in zone A](Other/Scotland_field_design.png){alt="Design of Scottish vegetation survey in zone A"}

![Design of Swiss trait sampling and vegetation surveys. The moss sample was collected in the centre of the plot.](Other/BDM_study_design.png)

![](Figures/LA_trait_values.png)

![](Figures/PlantHeight_trait_values.png)

![](Figures/SLA_trait_values.png)
